---
title: "Check CARseq results from SKCM analysis"
output:
  html_document:
  df_print: paged
---

# Load required library
```{r warning=FALSE, echo = TRUE, results = 'hide', message = FALSE}
library(data.table)
library(ggplot2)
library(ggrepel)

library(CARseq)
library(fgsea)
library(DESeq2)
theme_set(theme_bw())
```


# CARseq results

## Load results 

Read in gene expression, cell type proportion estimates, as well as covariates data

```{r}
res = readRDS("../results/SKCM_CARseq.rds")
lapply(res, function(x){if(is.vector(x)) length(x) else dim(x)})

res$p[1:2,]
res$padj[1:2,]

qval = apply(res$p, 2, get_qvalues_one_inflated)
dim(qval)
rownames(qval) = rownames(res$p)

colSums(res$padj < 0.1, na.rm=TRUE)
colSums(qval < 0.1, na.rm=TRUE)
```


## Read in gene annoations

```{r}
anno= fread("../data/gencode.v22.genes.txt")
dim(anno)
anno[1:2,]

table(rownames(res$p) %in% anno$geneId)
anno = anno[match(rownames(res$p), anno$geneId),]
dim(anno)
anno[1:2,]

anno$gene_name = anno$hgnc_symbol
table(anno$gene_name == "")
ww1 = which(anno$gene_name == "")
anno$gene_name[ww1] = anno$ensembl_gene_id[ww1]

gmtfile_reactome = file.path("../data/c2.cp.reactome.v7.1.symbols.gmt")
pathways_reactome = gmtPathways(gmtfile_reactome)
```

## valcano plot

```{r warning = FALSE, fig.width = 5, fig.height = 4}
qcut = 1e-6
logFC_cut = 5

cell_types = colnames(res$p)
res$shrunken_lfc[1:2,]
res$lfc[1:2,]

qval[1:2,]
table(rownames(qval) == rownames(res$shrunken_lfc))

for(ct1 in cell_types){
  cat(ct1, "\n")

  ct2 = paste0("1_vs_0:", ct1)
  df1 = data.frame(qvalue = qval[,ct1], logFC = res$shrunken_lfc[,ct2], 
                   orig_logFC = res$lfc[,ct2], 
                   gene = anno$gene_name, DE = rep(0, nrow(qval)))

  if(! any(df1$qvalue < 0.1, na.rm=TRUE)){
    cat("None of the genes have q-value < 0.1\n")
    next
  }
  
  df1[which(df1$qvalue < 0.1),]
  
  wwNA = which(is.na(df1$logFC))
  df1$logFC[wwNA] = df1$orig_logFC[wwNA]
  
  df1$DE[which(df1$qvalue < 0.1 & df1$logFC > 0)] = 1
  df1$DE[which(df1$qvalue < 0.1 & df1$logFC < 0)] = -1
  df1$gene[which(df1$DE == 0)] = ""

  df1$DE = as.factor(df1$DE)


  df1$qvalue[which(df1$qvalue < qcut)] = qcut
  df1$logFC[which(df1$logFC > logFC_cut)] = logFC_cut
  df1$logFC[which(df1$logFC < -logFC_cut)] = -logFC_cut

  g1 = ggplot(df1, aes(x=logFC, y=-log10(qvalue), col=DE)) +
    geom_point(size = 0.4) + xlim(-4.6, 4.6) + 
    scale_color_manual(values=c("#0072B2", "#CCCCCC", "#FC4E07")) +
    theme(legend.position = "none") + ggtitle(ct1)
  
  g2 = g1 + geom_text_repel(aes(label = gene), size = 3.5, 
                            min.segment.length = 0, 
                            box.padding = unit(0.25, "lines"),
                            point.padding = unit(0.25, "lines")) 
  print(g2)
}

```

## gene set enrichment anlaysis

```{r warning = FALSE}
fgsea_all = NULL
topP_list = list()

cell_types = colnames(res$p)

set.seed(1234)

for(ct1 in cell_types){
  cat(ct1, "\n")
  stats = -log10(res$p[,ct1])
  names(stats) = anno$hgnc_symbol
  # deduplicate by taking the one with highest -log10 p-value
  stats = na.omit(stats[unique(names(stats))])
  
  fgseaRes = fgseaMultilevel(pathways_reactome, stats, minSize=10, 
                             maxSize=1000, gseaParam = 0, eps = 0)
  
  # genes that are enriched in small p-values
  fgseaResPositive = fgseaRes[fgseaRes$NES > 0, ]
  fgseaResPositive$pathway = gsub("^REACTOME_", "", 
                                  fgseaResPositive$pathway, perl=TRUE)
  fgseaResPositive = fgseaResPositive[order(fgseaResPositive$padj),]
  
  fgseaResPositive$cell_type = rep(ct1, nrow(fgseaResPositive))
  fgsea_all = rbind(fgsea_all, fgseaResPositive)
  
  topPathways = fgseaResPositive[, c("pathway", "padj", "NES", "cell_type")]
  w2kp = which(topPathways$padj < 0.01)
  
  if(length(w2kp) == 0){ 
    next
  } else if(length(w2kp) > 10){ 
    w2kp = w2kp[1:10]
  }
  
  topPathways = topPathways[w2kp,]
  topPathways = topPathways[order(topPathways$padj),]
  
  topPathways$pathway = substr(topPathways$pathway, 1,60)

  print(topPathways[,-c("cell_type")])
  cat("\n")
  
  if(nrow(topPathways) > 0){
    topP_list[[ct1]] = topPathways
  }
  
}
```

## summarize GSEA results
```{r}
sapply(topP_list, nrow)
cbp1 = c("#D55E00","#E69F00", "#56B4E9", "#009E73",
          "#F0E442", "#999999", "#0072B2",  "#CC79A7")
names(cbp1) = cell_types

gp1 <- function(topPathways, color1){
  ggplot(data=topPathways, aes(x=pathway, y=-log10(padj), fill=cell_type)) + 
    geom_bar(stat="identity") + coord_flip() + 
    scale_x_discrete(limits = rev(topPathways$pathway)) + 
    scale_fill_manual(values=color1)
}
```


### CAFs
```{r fig.width = 9, fig.height = 3}
ct1 = "CAFs"
gp1(topP_list[[ct1]], cbp1[ct1])
```

### CD4_Tcells
```{r fig.width = 9, fig.height = 2.1}
ct1 = "CD4_Tcells"
gp1(topP_list[[ct1]], cbp1[ct1])
```

### Endothelial
```{r fig.width = 10, fig.height = 3}
ct1 = "Endothelial"
gp1(topP_list[[ct1]], cbp1[ct1])
```

### Macrophages
```{r fig.width = 6, fig.height = 1}
ct1 = "Macrophages"
gp1(topP_list[[ct1]], cbp1[ct1])
```

### NKcells
```{r fig.width = 9, fig.height = 3}
ct1 = "NKcells"
gp1(topP_list[[ct1]], cbp1[ct1])
```


# DESeq2 results

## Load results 

Read in gene expression, cell type proportion estimates, as well as covariates data

```{r}
resD = readRDS("../results/SKCM_DESeq2.rds")
dim(resD)
resD[1:2,]

table(rownames(resD) == anno$geneId)
resD$gene = anno$gene_name

resD$DE = rep(0, nrow(resD))
resD$DE[which(resD$padj < 0.1 & resD$log2FoldChange > 0)] = 1
resD$DE[which(resD$padj < 0.1 & resD$log2FoldChange < 0)] = -1
resD$gene[which(resD$DE == 0)] = ""
resD$DE = as.factor(resD$DE)

table(resD$padj < 0.1, resD$DE, useNA = "ifany")

dim(resD)
resD[1:2,]
```

## valcano plot
```{r warning = FALSE, fig.width = 5, fig.height = 4}
resD.df = as.data.frame(resD)
g1 = ggplot(resD.df, aes(x=log2FoldChange, y=-log10(padj), col=DE)) +
  geom_point(size = 0.4) + xlim(-4.6, 4.6) + 
  scale_color_manual(values=c("#0072B2", "#CCCCCC", "#FC4E07")) +
  theme(legend.position = "none") + 
  geom_text_repel(aes(label = gene), size = 3.5, 
                  min.segment.length = 0, 
                  box.padding = unit(0.25, "lines"),
                  point.padding = unit(0.25, "lines")) 
g1
```

## gene set enrichment anlaysis

```{r warning = FALSE}
stats = -log10(resD$pvalue)
names(stats) = anno$hgnc_symbol
stats = na.omit(stats[unique(names(stats))])

fgseaRes = fgseaMultilevel(pathways_reactome, stats, minSize=10, 
                           maxSize=1000, gseaParam = 0, eps = 0)

# genes that are enriched in small p-values
fgseaResPositive = fgseaRes[fgseaRes$NES > 0, ]
fgseaResPositive$pathway = gsub("^REACTOME_", "", 
                                fgseaResPositive$pathway, perl=TRUE)

topPathways = fgseaResPositive[, c("pathway", "padj", "NES")]
topPathways = topPathways[order(topPathways$padj),]

topPathways[1:10,]
```


```{r}
gc()
sessionInfo()
```

